# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r setup}
# clear all variables
rm(list = ls(all = TRUE))
```

### Load the data (i.e. read.csv())
Read CSV into R
```{r load_data}
step_data <- read.csv(unz("activity.zip", "activity.csv")
                      , header = TRUE 
                      ,sep = ",")

names(step_data)
nrow(step_data)
```

### Process/transform the data (if necessary) into a format suitable for your analysis


## What is mean total number of steps taken per day?


### Make a histogram of the total number of steps taken each day
```{r hist_total_num_steps_daily}
# sum steps by date
step_data.sum <- aggregate(step_data$steps ~ step_data$date
                           , data = step_data
                           , FUN = sum)
colnames(step_data.sum) <- c("date", "steps")
# LL:  step_data.sum$steps = step_data.sum[[2]] = step_data.sum[,2]
hist(step_data.sum$steps
     , main = "Total Number of Steps Taken Each Day"
     , xlab = "Steps"
     , labels = TRUE)

```


### Calculate and report the mean and median total number of steps taken per day
```{r mean_median_stats}
step_data.mean <- aggregate(step_data$steps ~ step_data$date
                            , data = step_data
                            , FUN = mean)
colnames(step_data.mean) <- c("date", "mean")

step_data.median <- aggregate(step_data$steps ~ step_data$date
                              , data = step_data
                              , FUN = median)
colnames(step_data.median) <- c("date", "median")

step_data.mean.median <- merge(step_data.mean, step_data.median)

step_data_mean <- mean(step_data$steps, na.rm=TRUE)
step_data_median <- median(step_data$steps, na.rm=TRUE)

# sanity check
summary(step_data)
```

Ignoring NA values of the average number of steps for that given interval results in the mean of `r step_data_mean` and a median of `r step_data_median`.


```{r results="asis"}
library(xtable)
xt <- xtable(step_data.mean.median)
print(xt, type="html")
```



## What is the average daily activity pattern?


### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r time_series_plot}
step_data.avg <- aggregate(step_data$steps ~ step_data$interval * step_data$date
                           , data = step_data
                           , FUN = mean
                           , na.rm = TRUE)
colnames(step_data.avg) <- c("interval", "date", "steps")
plot(x = step_data.avg$interval
     , y = step_data.avg$steps
     , xlab = "Interval"
     , ylab = "Number of steps"
     , type = "l")

# identify row with maximum number of steps
max_num_step_row <- step_data.avg$steps == max(step_data.avg$steps);
rec_max_num_steps <- step_data.avg[max_num_step_row,];

```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
The 5-minute interval, on average across all the days, which contains the maximum number of steps is `r rec_max_num_steps$interval` [on `r rec_max_num_steps$date` for `r rec_max_num_steps$steps` steps]


## Imputing missing values


### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r complete_recs}
complete_recs <- complete.cases(step_data)
```
The total number of incomplete is records is `r sum(!complete_recs)` [`r sum(!complete_recs)/nrow(step_data) *100`% incomplete]


### Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r fill_incomplete}
# fill NA records with average for that interval, ignoring NA's
step_data.avgperinterval <- aggregate(step_data$steps ~ step_data$interval
                                      , data = step_data
                                      , FUN = mean
                                      , na.rm = TRUE)
colnames(step_data.avgperinterval) <- c("interval", "steps")

getsteps <- function(p_interval) {
    # identify row for matching interval
    row <- which(step_data.avgperinterval$interval == p_interval)
    result <- step_data.avgperinterval$steps[row]
    
    return (ifelse( is.na(result)
                    , 0
                    , result))
    
    }

```


### Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r dup_dataset}
step_data_copy <- step_data;
step_data_copy$steps <- ifelse(is.na(step_data_copy$steps)
                               , getsteps(step_data_copy$interval)    
                               , step_data_copy$steps)

```


### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r hist_with_filled_values}
# sum steps by date
step_data_copy.sum <- aggregate(step_data_copy$steps ~ step_data_copy$date
                                , data = step_data_copy
                                , FUN = sum)
colnames(step_data_copy.sum) <- c("date", "steps")

hist(step_data_copy.sum$steps
     , main = "Total Number of Steps Taken Each Day"
     , xlab = "Steps"
     , labels = TRUE)

step_data_copy_mean <- mean(step_data_copy$steps)
step_data_copy_median <- median(step_data_copy$steps)

```

Filling NA values with the average number of steps for that given interval results in the mean of `r step_data_copy_mean` and a median of `r step_data_copy_median`.


* The impact of filling in the missing data with the average number of steps for a given interval is
    + a lower mean
    + no change in median since there are quite a few zero values



## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r weekday_or_weekend}
step_data_copy$wend <- NULL
step_data_copy$wend <- as.factor(ifelse(weekdays(as.Date(step_data_copy$date)) 
                                        %in% c("Saturday","Sunday")
                                        , "weekend"
                                        , "weekday")
                                 )

# sanity check
#step_data_copy$dow <- NULL
#step_data_copy$dow <- as.factor(weekdays(as.Date(step_data_copy$date)))

```

### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r weekday_weekend_plot}
step_data_copy.avg <- aggregate(step_data_copy$steps 
                                ~ step_data_copy$interval * step_data_copy$date * step_data_copy$wend
                                , data = step_data_copy
                                , FUN = mean)
colnames(step_data_copy.avg) <- c("interval", "date", "wend", "steps")


t_weekdays <- step_data_copy.avg[step_data_copy.avg$wend %in% c("weekday"),]
t_weekends <- step_data_copy.avg[step_data_copy.avg$wend %in% c("weekend"),]

par(mfrow = c(2, 1))
plot(x = t_weekdays$interval
     , y = t_weekdays$steps
     , main = "weekday"
     , xlab = "Interval"
     , ylab = "Number of steps"
     , type = "l")
plot(x = t_weekends$interval
     , y = t_weekends$steps
     , main = "weekend"
     , xlab = "Interval"
     , ylab = "Number of steps"
     , type = "l")

```
